Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-25 paper/drinking.tex

Automated reasoning in the Web Ontology Language (OWL) can be employed
to automatically compute the generalization hierarchy underlying an
ontology as well as for verification of data consistency and complex
queries \cite{Hoehndorf2011incon, Hoehndorf2011models}.

Layered approach: although OWL has high expressivity, some things
cannot be expressed in OWL; some things should not be expressed in OWL
for specific applications due to tractability reasons

@@

All GO regulation terms use one of three relations, regulates,
negatively\_regulates and positively\_regulates, to link regulation
terms to process or quality terms. The regulates relations are defined
in terms of qualities: a regulatory process causes a change in
magnitude to some quality, which in turn has an effect on the
frequency, rate or duration of some other type of process. Effects
that results in increases and decreases use positively\_regulates and
negatively\_regulates respectively. \cite{} (cite the paper on
cross-products in GO, PMID:20152934) The existing ontology structure
would also support the addition of subclasses to distinguish, for
example, regulation of the rate of a process from regulation of its
duration or time of onset.

quick intro...

mention regulation relations...

regulation of X processes

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO\cite {Ashburner2000b},
the Cell Type Ontology\cite {Bard2005a}, (SO)\cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} (PATO:0000499) with the Gene Ontology term \textit{estrous
  cycle} (GO:0042698), whilst if such a term existed in a precomposed
ontology (for the MP:0009007 term from the Mammalian Phenotype) it
could be used to provide an equivalence statement between that class
and the above PATO-based description.

>>>>

Automated reasoning in the Web Ontology Language (OWL) can be employed
to automatically compute the generalization hierarchy underlying an
ontology as well as for verification of data consistency and complex
queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
efficient automated reasoners are available to process OWL ontologies
\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
to support even large ontologies by further reducing the expressivity
of OWL in order to enable polynomial-time inferences. In particular
the OWL EL profile was found to provide the expressivity required for
most biomedical ontologies \cite{el4, elvira}, and highly optimized
OWL EL reasoners are available or under development to support
reasoning over very large ontologies \cite{el4, cbreasoner}.

A high expressivity is required to accurately specify complex axioms
that constrain the domain under investigation, and languages with
higher expressivity than OWL are often required in the biomedical
domain to achieve this goal \cite{Hoehndorf2009sequences,
  rnao}. On the other hand, automated reasoning over large ontologies
and associated datasets benefits from languages with a low complexity
of inferences in which complex axioms cannot be formulated. Therefore,
we use a layered approach: we specify the meaning of terms

although OWL has high expressivity, some things
cannot be expressed in OWL; some things should not be expressed in OWL
for specific applications due to tractability reasons

@@

All GO regulation terms use one of three relations, {\bf regulates},
{\bf negatively\_regulates} and {\bf positively\_regulates}, to link
regulation terms to process or quality terms. The {\bf regulates}
relations are defined in terms of qualities: a regulatory process
causes a change in magnitude to some quality, which in turn has an
effect on the frequency, rate or duration of some other type of
process. Effects that results in increases and decreases use {\bf
  positively\_regulates} and {\bf negatively\_regulates} respectively
\cite{Mungall2010go}. The existing ontology structure would also
support the addition of subclasses to distinguish, for example,
regulation of the rate of a process from regulation of its duration or
time of onset.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO\cite {Ashburner2000b},
the Cell Type Ontology\cite {Bard2005a}, (SO)\cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for the {\tt MP:0009007} term from
the Mammalian Phenotype) it could be used to provide an equivalence
statement between that class and the above PATO-based description.
